Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-07 talk/pepm2011/paper.tex
rewrite the introduction to mention partial evaluation more

Boxing of primitive types means that dynamic languages need to be able to handle
all objects, even integers, floats, booleans etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
heap-structure is allocated for them, that contains the actual value. Boxing
primitive types can be very costly, because a lot of common operations,
particularly all arithmetic operations, have to produce a new box, in addition
to the actual computation they do. Because the boxes are allocated on the heap,
producing a lot of them puts pressure on the garbage collector.

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. This
approach is also the one taken by the PyPy project, which is an environment for
implementing dynamic programming languages. PyPy's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient virtual machine (VM) that also contains a just-in-time compiler.
This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}.

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by PyPy's tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects}\footnote{The terminology comes from \cite{rigo_representation-based_2004}},
which is described in Section~\ref{sec:virtuals}. The
goal of virtual objects is to remove allocations of temporary objects that have
a predictable lifetime and to optimize type dispatching in the process.

The basic approach of virtual objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:support} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.

@@

\begin{enumerate}
    \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
    \item XXX
\end{enumerate}

@@

\section{Formal Description of the Algorithm}
\label{sec:Algorithm}

>>>>

Boxing of primitive types is necessary because dynamic languages need to be able to handle
all objects, even integers, floats, booleans etc. in the same way as user-defined
instances. Thus those primitive types are usually \emph{boxed}, i.e. a small
heap-structure is allocated for them, that contains the actual value. Boxing
primitive types can be very costly, because a lot of common operations,
particularly all arithmetic operations, have to produce a new box, in addition
to the actual computation they do. Because the boxes are allocated on the heap,
producing a lot of them puts pressure on the garbage collector.

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of an interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which makes many optimizations that are usually
hard in a compiler simpler, such as register allocation.

The usage of a tracing JIT can remove the overhead of bytecode dispatch and that
of the interpreter data structures. In this paper we want to present an approach
how an optimization can be added to a tracing JIT that also removes some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. Our experimental platform is the PyPy project, which is an
environment for implementing dynamic programming languages. PyPy and tracing
JITs are described in more detail in Section~\ref{sec:Background}.
Section~\ref{sec:lifetimes} analyzes the problem to be solved more closely.

The most important technique we use to achieve to optimize traces is a
straightforward application of partial evaluation. The partial evaluation
performs a form of escape analysis \cite{XXX} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:virtuals}, a more formal description is
given in Section~\ref{sec:formal}.

The basic approach of static objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:support} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.

@@

\begin{enumerate}
    \item An efficient and effective algorithm for removing object allocations in a tracing JIT.
    \item A characterization of this algorithm as partial evaluation.
    \item A rigorous evaluation of this algorithm.
\end{enumerate}

@@

\section{Formal Description of the Algorithm}
\label{sec:formal}
